---
title: "Final Model Analysis and Interpretation"
subtitle: "ADS 503 Final Project"
author: "Vince Garcia"
format: pdf
---
Libraries
```{r setup}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
library(caret)
library(pROC)
library(scales)
library(glmnet)
library(ggplot2)
library(gbm)
library(PRROC) # Explicitly loaded for Precision-Recall evaluation metrics

set.seed(503)

```
Importing from Model Tuning
```{r rebuild-data}
#| warning: false
setwd(r"(C:\Users\User\ADS503_Team5\Raw Data)")
load("evaluated_models.RData")
#source: https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/load

```
Model Selection & Justification
```{r}
# Plot the visual justification (ROC)
cal_obj <- calibration(obs ~ LDA+ElasticNet+RF+GBM, data = calibration_df, cuts = 11)
plot(cal_obj, type = "b", main = "Probability Calibration Curve", auto.key = TRUE)
```
From this Proabability Calibration Curve, all models are trained on the same test data. As the data spreads out, each model improves its observed event percentage up. However, when the bin midpoint reaches 100, LDA takes a noticeable drop while GBM maintains the highest Observed Event Percentage. 
Highest overall performance: GBM
```{r}
#Performance metrics summarized
summary(res)
```
Highest Mean ROC: GBM
Highest Mean Sensitivity: Random Forest
Highest Mean Specificity: LDA

Among the 4 models tested, GBM has the highest ROC, which can be determined as the overall performance. While the ROC is a reliable measure of overall performance, the linear model, LDA does have the highest specificity while Random Forest has the highest sensitivity.

Important Variables for Winning Model
```{r}
#using var imp for the important variables
#source: https://r-statistics.co/caret-varImp-in-R.html
important_variables <- varImp(mod_gbm, scale = TRUE)
print(important_variables$importance)
```
Above are the weights of each variable in the GBM model. General Health remains the most critical variable to the model with an overall % score of 100, followed by HighBP (77.06%) and BMI (44.63%).


Visual Representation of Variable Importance of GBM
```{r}
#source: https://r-statistics.co/caret-varImp-in-R.html
ggplot(important_variables) +
  labs(title = "Variable importance from GBM",
       x = "Predictor", y = "Importance (0% to 100%)") +
  theme_minimal()
```
The distribution of importance shows that the leading indicators of diabetes are: General Health (100%), High BP (77.06%), BMI (44.63%), Age (30.15%), HighChol (23.91%), and DiffWalk (4.96%).

Confusion Matrix
```{r}
print(gbm_cm)
```

The final values for the GBM Model are:
ROC AUC           : 0.8302
Accuracy          : 0.7504 
Balanced Accuracy : 0.7504
Sensitivity       : 0.7748          
Specificity       : 0.7260 