---
title: "Model Evaluation"
subtitle: "ADS 503 Final Team Project - Diabetes Health Indicators (BRFSS 2015)"
author: "Vince Garcia"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
  echo: true
---
Libraries
```{r setup}
library(tidyverse)
library(caret)
library(recipes)
library(pROC)
library(scales)
library(doParallel)
library(glmnet)
library(ggplot2)
library(ranger)
library(gbm)
library(PRROC) # Explicitly loaded for Precision-Recall evaluation metrics

set.seed(503)
dm_palette <- c("No Diabetes" = "#4C9F70", "Diabetes" = "#D1495B")

```
Importing from Model Tuning
```{r rebuild-data}

setwd(r"(C:\Users\User\ADS503_Team5\Raw Data)")


load("tuned_models_and_data.RData")
#source: https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/load

```

ROC Curve Construction
```{r}
#Source: Robin (2026). Package 'pROC'. 
#Link:  https://xrobin.github.io/pROC/
prob_lda <- predict(mod_glm, x_test_processed, type = "prob")[["Diabetes"]] #gemini suggested [["Diabetes"]] subset when calling predict()
prob_enet <- predict(mod_glmnet, x_test_processed, type = "prob")[["Diabetes"]] #gemini suggested [["Diabetes"]] subset when calling predict()
prob_rf    <- predict(mod_rf, x_test_df, type = "prob")[["Diabetes"]]
#gemini suggested [["Diabetes"]] subset when calling predict()
prob_gbm    <- predict(mod_gbm, x_test_df, type = "prob")[["Diabetes"]]
#gemini suggested [["Diabetes"]] subset when calling predict()




roc_lda  <- roc(test_y, prob_lda, levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
roc_enet    <- roc(test_y, prob_enet, levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
roc_rf  <- roc(test_y, prob_rf, levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
roc_gbm     <- roc(test_y, prob_gbm, levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
```

Delong Test Results for model ROC Curves
```{r}
# Execute statistical comparison of holdout curves
#LDA Vs. Elastic Net
delong_result <- roc.test(roc_lda, roc_enet, method = "delong")
print(delong_result)
#Random Forest vs. GBM
delong_result2 <- roc.test(roc_rf, roc_gbm,method = "delong")
print(delong_result2)
#LDA versus GBM
delong_result3 <- roc.test(roc_lda, roc_gbm,method = "delong")
print(delong_result3)
#Random Forest versus LDA
delong_result4 <- roc.test(roc_rf, roc_lda,method = "delong")
print(delong_result4)
#Random Forest versus Elastic Net
delong_result5 <- roc.test(roc_rf, roc_enet,method = "delong")
print(delong_result5)
```
Judging by the results of the Delong's tests, most of the models remain statistically indistinguishable from each other except for LDA and GBM, where the p value of that test is 0.002433. GBM outperforms LDA significantly. 



ROC Comparison
```{r}
##Source: Robin (2026). Package 'pROC'. 
#Link:  https://xrobin.github.io/pROC/
# Build individual ROC objects first
r_lda   <- roc(test_y, prob_lda,   levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
r_enet <- roc(test_y, prob_enet, levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
r_rf    <- roc(test_y, prob_rf,    levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
r_gbm   <- roc(test_y, prob_gbm,   levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)

# Put them in a named list
#suggested by Gemini:
roc_list <- list(LDA = r_lda, ElasticNet = r_enet, RandomForest = r_rf, GBM = r_gbm)

# Plot all curves simultaneously with ggplot2
#source : https://www.rdocumentation.org/packages/ggROC/versions/1.0/topics/ggroc
ggroc(roc_list, aes = "color", linewidth = 1, alpha = 0.9) +
  theme_minimal() +
  labs(title = "ROC Comparison Across Models", color = "Algorithm", x = "1- Specificity (%)", y = "Sensitivity (&)")
```
Probability Calibration Assessment
```{r}
# Create a dedicated dataframe for calibration
calibration_df <- data.frame(
  obs        = test_y,
  LDA        = prob_lda,
  ElasticNet = prob_enet,
  RF         = prob_rf,
  GBM        = prob_gbm
) #suggested by Gemini

#source: Kuhn (2026). Package 'caret': 'calibration'". 
#https://cran.r-project.org/web/packages/caret/caret.pdf

#source for auto.key
#https://stackoverflow.com/questions/10828016/lattice-auto-key-how-to-adjust-lines-and-points
cal_obj <- calibration(obs ~ LDA+ElasticNet+RF+GBM, data = calibration_df, cuts = 11)
plot(cal_obj, type = "b", main = "Probability Calibration Curve", auto.key = TRUE)

```
Model summaries
```{r}
#Creating a list of the models loaded in
#source: https://www.w3schools.com/r/r_lists.asp

models <- list(
  LDA     = mod_glm,
  ElasticNet   = mod_glmnet,
  RandomForest = mod_rf,
  GBM          = mod_gbm
)
#Using resamples to give a summary of the models
#https://www.rdocumentation.org/packages/caret/versions/6.0-92/topics/resamples
res <- resamples(models)
summary(res)
```
Highest Mean ROC: GBM
Highest Mean Sensitivity: Random Forest
Highest Mean Specificity: LDA

Model Comparison
```{r}
bestTunes <- list(
  LDA     = mod_glm$bestTune,
  ElasticNet   = mod_glmnet$bestTune,
  RandomForest = mod_rf$bestTune,
  GBM          = mod_gbm$bestTune
)
print(bestTunes)
#Using diff to give a summary of the resamples
#https://www.rdocumentation.org/packages/caret/versions/6.0-92/topics/resamples
resamples_diff <- diff(res) #gemini suggested using diff() on the entire res (resample) variable
summary(resamples_diff) #gemini suggested the use of summary() for the final results
```

After evaluating the summary of differences between models. There are multiple conclusions that may be drawn. 
ROC:

LDA vs. GBM : p value is 0.03812. Difference is -2.538e-03. This is statistically significant (P<0.05)  and despite that small difference in ROC, GBM does outperform LDA. 

Elastic Net vs. GBM: p-value is 0.04356. Difference is -2.498e-03. This difference in overall performace is statistically significant (P<0.05)  and GBM does outperform elastic net. 

Sensitivity:
GBM vs. LDA: p-value is 0.006523. Difference is  -0.008381. The result is statistically significant (P<0.05)  and despite the small difference, GBM does outperform LDA in sensitivity. 

Random Forest vs. LDA: p-value is 0.001598 Difference is -0.022775 The result is statistically significant (P<0.05) and despite the small difference, Random Forest does outperform LDA in sensitivity. 

Random Forest vs. Elastic Net: p-value is 0.003055  Difference is -0.023164   The result is statistically significant (P<0.05)  and despite the small difference, Random Forest does outperform Elastic Net in sensitivity. 

GBM vs. Elastic Net: p-value is 0.019979     Difference is -0.008770   The result is statistically significant (P<0.05) and despite the small difference, GBM does outperform Elastic Net in sensitivity. 

GBM vs. Random Forest: p-value is 0.001068     Difference is -0.0129078   The result is statistically significant (P<0.05) and despite the small difference, GBM does outperform Random Forest in sensitivity. 

Specificity:
Random Forest vs. LDA: p-value is 0.01421 . Difference is 0.0185307. The result is statistically significant (P<0.05) and despite the small difference, LDA does outperform Random Forest in specificity 

Random Forest vs. Elastic Net: p-value is 0.01206 . Difference is 0.0183185 The result is statistically significant (P<0.05) and despite the small difference, Elastic Net does outperform Random Forest in specificity 



