---
title: "Model Tuning/Evaluation"
subtitle: "ADS 503 Final Team Project — Diabetes Health Indicators (BRFSS 2015)"
author: "Malik Tisani"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
  echo: true
---

# Model Evaluation

> Builds on the preprocessing notebook (`train_processed` / `test_processed`,
> seed 503, 80/20 stratified split). Baseline logistic regression is the floor
> (test Accuracy ≈ 0.747, AUC ≈ 0.827). We fit the candidate models under one
> consistent cross-validation harness, compare them, and select a final model,
> prioritizing ROC AUC and Balanced Accuracy.
>
> Textbook map (Kuhn & Johnson, *Applied Predictive Modeling*): resampling &
> tuning → Ch. 4; classification metrics & ROC → Ch. 11; importance → Ch. 18.

```{r setup}
library(tidyverse)
library(caret)
library(recipes)
library(pROC)
library(scales)

# Model engines (install once if needed):
# install.packages(c("glmnet", "ranger", "gbm"))
library(glmnet)
library(ranger)
library(gbm)

set.seed(503)
dm_palette <- c("No Diabetes" = "#4C9F70", "Diabetes" = "#D1495B")
```

## Rebuild processed data (standalone)

Reproduces the split and recipe so this file renders on its own. (the
`if (!exists(...))` guard does that automatically).

```{r rebuild-data}
if (!exists("train_processed") || !exists("test_processed")) {

  df <- read_csv("diabetes_binary_5050split_health_indicators_BRFSS2015.csv") |>
    mutate(Diabetes_binary = factor(
      Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")
    ))

  set.seed(503)
  split_index <- createDataPartition(df$Diabetes_binary, p = 0.80, list = FALSE)
  train_data  <- df[split_index, ]
  test_data   <- df[-split_index, ]

  diabetes_recipe <- recipe(Diabetes_binary ~ ., data = train_data) |>
    step_zv(all_predictors()) |>
    step_normalize(all_numeric_predictors())

  diabetes_prep   <- prep(diabetes_recipe, training = train_data)
  train_processed <- bake(diabetes_prep, new_data = train_data)
  test_processed  <- bake(diabetes_prep, new_data = test_data)
}
```



## caret-safe target coding


```{r relabel}
recode_target <- function(d) {
  d |>
    mutate(Diabetes_binary = factor(
      ifelse(as.character(Diabetes_binary) == "Diabetes", "Diabetes", "NoDiabetes"),
      levels = c("Diabetes", "NoDiabetes")
    ))
}

train_cv <- recode_target(train_processed)
test_cv  <- recode_target(test_processed)

positive_class <- "Diabetes"

# Safety checks — these should print TRUE / no spaces before modeling
stopifnot(identical(levels(train_cv$Diabetes_binary), c("Diabetes", "NoDiabetes")))
stopifnot(!anyNA(train_cv))
```

## Shared resampling harness

All models train on the **same folds** (via `index`), so `resamples()` is a valid
paired comparison (Ch. 4). `savePredictions = "final"` keeps the out-of-fold
predictions for threshold tuning.

```{r ctrl}
set.seed(503)
cv_index <- createFolds(train_cv$Diabetes_binary, k = 5, returnTrain = TRUE)

ctrl <- trainControl(
  method          = "cv",
  index           = cv_index,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  verboseIter     = FALSE
)
```

## Fit the candidate models


```{r fit-glm}
set.seed(503)
mod_glm <- train(
  Diabetes_binary ~ ., data = train_cv,
  method = "glm", family = "binomial",
  metric = "ROC", trControl = ctrl
)
```

```{r fit-glmnet}
set.seed(503)
mod_glmnet <- train(
  Diabetes_binary ~ ., data = train_cv,
  method = "glmnet",
  metric = "ROC", trControl = ctrl,
  tuneLength = 10
)
```

```{r fit-rf}
# ranger uses all cores; impurity importance is fast and fine for varImp
mod_rf <- train(
  Diabetes_binary ~ ., data = train_cv,
  method = "ranger",
  metric = "ROC", trControl = ctrl,
  tuneLength = 3,
  num.trees = 300,
  importance = "impurity",
  verbose = FALSE
)
```

```{r fit-gbm}
# Stochastic gradient boosting (stable caret engine; no xgboost version issues)
gbm_grid <- expand.grid(
  n.trees           = c(150, 300),
  interaction.depth = c(2, 4),
  shrinkage         = 0.1,
  n.minobsinnode    = 10
)

set.seed(503)
mod_gbm <- train(
  Diabetes_binary ~ ., data = train_cv,
  method = "gbm",
  metric = "ROC", trControl = ctrl,
  tuneGrid = gbm_grid,
  verbose = FALSE
)
```

```{r model-list}
models <- list(
  Logistic     = mod_glm,
  ElasticNet   = mod_glmnet,
  RandomForest = mod_rf,
  GBM          = mod_gbm
)
```

## Cross-validated comparison

```{r resamples}
res <- resamples(models)
summary(res)

bwplot(res, metric = "ROC")
dotplot(res, metric = c("ROC", "Sens", "Spec"))

summary(diff(res))   # are the gaps real or resampling noise?
```

## Test-set performance

```{r test-metrics}
class_results <- imap_dfr(models, function(m, nm) {
  cls  <- predict(m, test_cv)
  cm   <- confusionMatrix(cls, test_cv$Diabetes_binary, positive = positive_class)
  prob <- predict(m, test_cv, type = "prob")[[positive_class]]
  roc_obj <- roc(response = test_cv$Diabetes_binary, predictor = prob,
                 levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
  tibble(
    Model       = nm,
    Accuracy    = cm$overall["Accuracy"],
    Kappa       = cm$overall["Kappa"],
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"],
    BalancedAcc = cm$byClass["Balanced Accuracy"],
    AUC         = as.numeric(auc(roc_obj))
  )
}) |>
  arrange(desc(AUC))

knitr::kable(class_results, digits = 3,
             caption = "Test-set performance across candidate models")
```

## Best model — confusion matrix & ROC

```{r best-model}
best_name  <- class_results$Model[1]
best_model <- models[[best_name]]

confusionMatrix(predict(best_model, test_cv), test_cv$Diabetes_binary,
                positive = positive_class)
```

```{r roc-curves}
roc_list <- imap(models, function(m, nm) {
  prob <- predict(m, test_cv, type = "prob")[[positive_class]]
  roc(response = test_cv$Diabetes_binary, predictor = prob,
      levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
})

ggroc(roc_list, legacy.axes = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey60") +
  labs(title = "ROC curves on the test set",
       x = "1 - Specificity", y = "Sensitivity", color = "Model")
```

## Threshold tuning for the chosen model


```{r threshold-tuning}
oof <- best_model$pred
bt  <- best_model$bestTune
oof <- merge(oof, bt)       
stopifnot(nrow(oof) > 0)

roc_oof  <- roc(response = oof$obs, predictor = oof[[positive_class]],
                levels = c("NoDiabetes", "Diabetes"), quiet = TRUE)
best_thr <- as.numeric(coords(roc_oof, "best", best.method = "youden",
                              ret = "threshold")$threshold[1])  
cat("Chosen probability threshold:", round(best_thr, 3), "\n")

test_prob       <- predict(best_model, test_cv, type = "prob")[[positive_class]]
test_pred_tuned <- factor(ifelse(test_prob >= best_thr, "Diabetes", "NoDiabetes"),
                          levels = c("Diabetes", "NoDiabetes"))

cm_default <- confusionMatrix(predict(best_model, test_cv),
                              test_cv$Diabetes_binary, positive = positive_class)
cm_tuned   <- confusionMatrix(test_pred_tuned,
                              test_cv$Diabetes_binary, positive = positive_class)

tibble(
  Threshold   = c("Default (0.50)", paste0("Tuned (", round(best_thr, 2), ")")),
  Sensitivity = c(cm_default$byClass["Sensitivity"], cm_tuned$byClass["Sensitivity"]),
  Specificity = c(cm_default$byClass["Specificity"], cm_tuned$byClass["Specificity"]),
  BalancedAcc = c(cm_default$byClass["Balanced Accuracy"], cm_tuned$byClass["Balanced Accuracy"])
) |>
  knitr::kable(digits = 3, caption = "Default vs. tuned threshold (test set)")
```

## Predictor importance

```{r var-imp}
vi <- varImp(best_model)
plot(vi, top = 15)
```

## Final model selection & discussion

We selected the stochastic gradient boosting model (GBM) as the final model. Across the five resampling folds it produced the highest mean cross validated ROC (0.830), exceeding random forest (0.826), elastic net (0.824), and logistic regression (0.824). To confirm that this ranking reflected a real performance difference rather than fold to fold variation, we examined the paired resampling differences. The boosting model significantly outperformed both linear models on ROC (Bonferroni adjusted p = 0.004 versus logistic regression and p = 0.004 versus elastic net) and also improved significantly on random forest (p = 0.031). In contrast, the two linear models were statistically indistinguishable from one another (p = 1.000), and random forest did not separate cleanly from the linear models (p > 0.07). The held out test set reproduced this ordering: the boosting model achieved the strongest test ROC AUC (0.834), Balanced Accuracy (0.754), and Cohen's Kappa (0.508), placing it ahead of every competing model on the metrics the team prioritized.

The practical margin over the logistic baseline is modest, roughly 0.007 in AUC, which warrants an honest framing for a stakeholder audience. The boosting model is the statistically and empirically superior choice, and it remains interpretable through its variable importance profile, where general health status, high blood pressure, and body mass index dominate. This profile aligns closely with the exploratory analysis, which identified the same predictors as the strongest univariate signals, and it lends the model clinical face validity. We therefore recommend the boosting model as the final solution while noting that penalized logistic regression performs within a narrow band on every metric and would be a defensible alternative where maximal transparency outweighs a small gain in discrimination.

The diabetic class carries a higher cost when missed, so we did not rely on the default 0.50 decision threshold. Following the alternate cutoff procedure, we selected the probability threshold that maximizes Youden's J index, computed on the out of fold resampling predictions so that the test set played no role in the choice. The resulting threshold of 0.476 raised test sensitivity from 0.793 to 0.814 at a controlled cost to specificity (0.715 to 0.696), improving the model's ability to identify true diabetic cases while holding overall balanced accuracy steady.

**Limitations.**

Two limitations qualify these results. First, the modeling data used the balanced fifty fifty file, which means the test set is also balanced rather than reflecting the approximately fourteen percent diabetes prevalence observed in the full population. Performance measures that depend on prevalence, particularly positive predictive value, are therefore optimistic relative to a real screening deployment, and a validation pass on a holdout drawn from the full distribution would give a more realistic estimate of operational performance. Second, several data quality items surfaced during the exploratory analysis were not addressed in the preprocessing recipe, including implausible body mass index values reaching 98 and the strong zero inflation in the self reported mental and physical health day counts. Incorporating these adjustments, along with calibration analysis and cost sensitive thresholding, represents a natural extension of this work.
